Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-07-04 workflow/workflow.Rmd
updated Slingshot section, per Sandrine's comments.

---
title: "Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
  - name: Kelly N. Street
    affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
  - name: Sandrine Dudoit
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley; Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA

@@

To infer lineages and pseudotimes, we use the package `slingshot`, with input the 4-dimensional MDS of the low-dimensional matrix `W`. We found that the slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_ from 2 to 5). Here, we use the unsupervised version of `slingshot`, where we only provide the identity of the start cluster but not of the end clusters. This implementation recovers the lineages found in the published work (Figures \@ref(fig:curves) and \@ref(fig:tree)).

[SD: Refer to 3D version, used to create Figure 2.]

@@

lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1")
lineages <- getCurves(lineages)
```

@@

```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
```

@@

@@

In the workflow, we recover a reasonable ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. `getLineages` has the option for the user to provide known end cluster(s). Here is the code to use `slingshot` in a supervised setting.

```{r slingshotsupervised, eval=FALSE}
lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
                         end.clus = c("c3", "c6"))
lineagees <- getCurves(lineages)
pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])

@@

FP, DR, KNS, and EP performed the data analaysis and wrote the code portions of the workflow. FP and SD wrote the text portion of the workflow, with contributions from the other three authors. DR, EP, and SD supervised the research.

@@

DR, KNS, EP, and SD were supported by the National Institutes of Health BRAIN Initiative (U01 MH105979, PI: John Ngai). KNS was supported by a training grant from the National Human Genome Research Institute (T32000047).

>>>>

---
title: "Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
  - name: Kelly Street
    affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
  - name: Sandrine Dudoit
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley; Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA

@@

To infer lineages and pseudotimes, we use the package `slingshot`, with input the 4-dimensional MDS of the low-dimensional matrix `W`. We found that the slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_ from 2 to 5). Here, we use the unsupervised version of `slingshot`, where we only provide the identity of the start cluster but not of the end clusters.

@@

lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1")
```

@@

Before fitting the simultaneous principal curves, we check the structure of the lineages by plotting the MST on the clusters. This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)). The `slingshot` package also includes methods for 3-dimensional visualizations similar to \@ref(fig:stemcelldiff), using the `plot3d` function from the package `rgl`.

@@

Having found the global lineage structure, we now construct a set of smooth, branching curves in order to infer the pseudotime variables. Simultaneous principal curves are constructed from the individual cells along each lineage, rather than the cell clusters. This makes them more stable and better suited for assigning cells to lineages. The final curves are shown in \@ref(fig:curves).

```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
lineages <- getCurves(lineages)

pairs(lineages, type='curves', col = pal[primaryClusterNamed(ceObj)])
```

@@

In the workflow, we recover a reasonable ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. `getLineages` has the option for the user to provide known end cluster(s). Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c8` and `c6` represent terminal cell fates.

```{r slingshotsupervised, eval=FALSE}
lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
                         end.clus = c("c3", "c6"))
lineagees <- getCurves(lineages)
pairs(lineages, type='curves', col = pal[primaryClusterNamed(ceObj)])
pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)], show.constraints = TRUE)

@@

FP, DR, KS, and EP performed the data analaysis and wrote the code portions of the workflow. FP and SD wrote the text portion of the workflow, with contributions from the other three authors. DR, EP, and SD supervised the research.

@@

DR, KS, EP, and SD were supported by the National Institutes of Health BRAIN Initiative (U01 MH105979, PI: John Ngai). KS was supported by a training grant from the National Human Genome Research Institute (T32000047).
